### 2010-03-29 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: fix spacing of operators used in text.

However, this new class of objects cannot be defined by mere
induction. In the case of vectors, for instance, we have to define the
whole \emph{family} of vector in one go: vectors of all size needs to
be defined at the same time. Where simply-typed languages have
inductive types, in dependently typed programming we the basic grammar of
data-types should be that of inductive families of types. To this end, we
rely on \emph{indexing}. We are going to transform the $\Desc$ universe into
$\IDesc$, the universe of indexed descriptions. We express inductive families
by the $\V{I} \To \IDesc{\V{I}}$ type.  \note{pwm: Vec might not be the best
motivating example here, since it doesn't appear in the paper (more on that
later). Come to think of it, this is entirely too vague without spelling out
a concrete example.}

@@

Before moving on indexed descriptions, we have to carry out some
maintenance work on descriptions. We presented $\Desc$ as the grammar
of inductive types. Hence, the codes closely follow this grammar. In
the following, we adopt an alternative presentation. With $\DSigma$,
we are actually \emph{quoting} a standard type-former, namely

@@

In the alternative presentation, we go further and present all our
codes as quotations of standard type-formers. This presentation is
shown in Figure~\ref{fig:type-former-desc}. The reader will notice
that we replace $\DUnit$ by a more general $\DConst$ code. Whereas
$\DUnit$ was interpreted as the unit set, $\DConst{\V{X}}$ is interpreted
as $\V{X}$, for any $\Bhab{\V{X}}{\Set}$. Extensionally, $\DConst{\V{X}}$ and
$\DSigma{\V{X}}{\DUnit}$ are equivalent. However, $\DConst$ is more
succinct. More importantly, $\DConst$ is \emph{first-order}, unlike
its equivalent encoding. From a definitional perspective, we are
giving more opportunities to the type-system, hence reducing the
burden on the programmer. For the same reason, we introduce
$\DProd{}{}$ that overlaps with $\DSigma{}{}$.

This reorganisation is strictly equivalent to the previous one
(Fig.~\ref{fig:hindx_desc}). Just as the previous version, it is also
self-descriptive. We refer the reader to the companion technical
report for details. In this finer-grained presentation, we can define
$\DIndx$ and $\DHindx$ as follow:

@@

\[\stk{
\NatD : \Desc \\
\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{} ]})}
                     {[ \DUnit \quad \DId ]}
}\]

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where                                      \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DId            & \Desc                                    \\
    \DConst{}       & \Set \To \Desc                           \\
    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\Desc} \Desc          \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc \\
    \DPi            & \PI{\V{S}}{\Set} \PIS{\V{S} \To \Desc} \Desc 
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\descop{\DId}{\V{X}}          &  \V{X}                                           \\
\descop{\DConst{\V{Z}}}{\V{X}}    &  \V{Z}                                           \\
\descop{\DProd{\V{D}}{\V{D'}}}{\V{X}} &  \TIMES{\descop{\V{D}}{\V{X}}}{\descop{\V{D}\V{'}}{\V{X}}}       \\
\descop{\DSigma{\V{S}}{\V{D}}}{\V{X}} &  \SIGMA{\V{s}}{\V{S}} \descop{\V{D}\: \V{s}}{\V{X}}          \\
\descop{\DPi{\V{S}}{\V{D}}}{\V{X}}    &  \PI{\V{s}}{\V{S}} \descop{\V{D}\: \V{s}}{\V{X}}            
\end{array}
}
\end{array}
}\]

@@

In the previous section, we have presented the $\Desc$ universe as the
grammar of functors in the category $\Set$. We have seen how to code
inductive types in this setting. To describe an inductive family
indexed by $\Bhab{\V{I}}{\Set}$, we use endofunctors on the category
$\Set^{\V{I}}$. We call these \emph{indexed functors}. \note{Predicate
  transformer anyone?} In this setting, $\V{I} \To \IDesc{\V{I}}$ is our
grammar for describing these functors. Hence, $\IDesc{}$ and its
interpretation have the following types:

\[\stk{
\data \IDesc{} (\Bhab{\V{I}}{\Set}) : \Set \where \\
\;\;\ldots \\
\\
\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{\V{I}} \To (\V{I} \To \Set) \To \Set    \\
\ldots
}\]

@@

However, we still have to define the actual grammar. This can be
achieved by a minor evolution of $\Desc$
(Fig.~\ref{fig:type-former-desc}). Indeed, the functorial nature of
$\Desc$ is entirely captured by the $\DId$ code. We get an indexed
functor by turning $\DId$ into an index-dependent $\DVar$ code. The
resulting code of $\IDesc$ is presented in Figure~\ref{fig:idesc}.

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \IDesc{} (\Bhab{\V{I}}{\Set}) : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DVar{}         & \V{I} \To \IDesc{\V{I}}                                   \\
    \DConst{}       & \Set \To \IDesc{\V{I}}                                \\
    \DProd{}{}      & \PI{\V{\V{D}}, \V{\V{D'}}}{\IDesc{\V{I}}} \IDesc{\V{I}}           \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}}  \\
    \DPi            & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}} 
\end{array}
}
\vspace{0.2in}
\\
\stk{
\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{\V{I}} \To \Set) \To \Set                  \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{\V{i}}}{\V{I}}{\V{P}}      &  \V{P}~\V{i}                                                 \\
\idescop{\DConst{\V{X}}}{\V{I}}{\V{P}}    &  \V{X}                                                   \\
\idescop{\DProd{\V{D}}{\V{D'}}}{\V{I}}{\V{P}} &  \TIMES{\idescop{\V{D}}{\V{I}}{\V{P}}}{\idescop{\V{D'}}{\V{I}}{\V{P}}}       \\
\idescop{\DSigma{\V{S}}{\V{D}}}{\V{I}}{\V{P}} &  \SIGMA{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}                    \\
\idescop{\DPi{\V{S}}{\V{D}}}{\V{I}}{\V{P}}    &  \PI{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}            
\end{array}
}
\end{array}
}\]

@@

In the case of vectors, for instance, for the index $\NatZero$, we
would only propose the constructor $\ListNil$. Similarly, for
$\NatSuc{n}$, we only propose the constructor $\ListCons{}{}$. Hence,
we obtain the following definition:

@@

In order to gain some intuition with $\IDesc$, let us re-implement the
pattern functor of natural numbers. We refer the reader to
Section~\ref{sec:desc-examples} for its incarnation in the universe of
descriptions. The code is the following:

\[\stk{
\NatD : \IDesc{\Unit} \\
\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{} ]})}
                     {[ \DConst{\Unit} \quad (\DVar{\Void}) ]}
}\]

@@

A mandatory exercise consists in describing $\IDesc$ in itself. With
the experience gained in Section~\ref{sec:desc-levitate}, this
levitation does not pose any issue:

\[\stk{
\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescD~\V{I} \mapsto \DSigma{}{} & (\EnumT \red{[} \DVar{},
                                          \DConst{},
                                          \DProd{}{},
                                          \DSigma{}{}, 
                                          \DPi{}{} \red{]}) \\
                              & \bigRedBracket{\begin{array}{l}
                                      \DConst{\V{I}}                  \\
                                      \DConst{\Set}               \\
                                      \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                      \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
                                      \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
                                   \end{array}
                             }
\end{array}
}\]

Note that this description also lives in $\IDesc{\Unit}$. Indeed,
for any $\V{I}$, $\IDesc~\V{I}$ is merely an inductive type. As for the
$\Desc$ levitation, the finite function space $[ \ldots ]$ elaborates into a
special purpose $\F{switchID}$ operator.

@@

In our universe of descriptions, this data-type is represented by a
tagged indexed description. We use the index to carry the type: the
resulting description is indexed by $\Ty$. We observe that some
constructors are ``polymorph'', namely $\Econd{}{}{}$ and $\Eval{}$. On the
other hand, the $\Eplus$ and $\Ele$ constructors are
index-dependent. $\Eplus$ is defined if and only if the result type --
the index -- is $\Enat$, whereas $\Ele$ is defined if and only if the
index is $\Ebool$. The actual code precisely follows this intuition,
as shown in Figure~\ref{fig:hexpr-full}. For brevity, we use an
informal $\case{\ldots}{\ldots}$ notation, simulating a definition by
pattern-matching. Formally, this corresponds to a call to the
$\switch{}{}{}{}$ eliminator.

@@

Another common dependent data-type is $\Fin$, the finite sets of
numbers. $\Fin$ is a typical example of an inductive family using
\emph{constrained constructors}. The constructors $\FinZero$ and
$\FinSuc$ are only defined for an index distinct from $\NatZero$:

\[
\stk{
\data \Fin{} : \PI{\V{n}}{\Nat} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FinZero_{\V{n}}      & \Fin{(\NatSuc{\V{n}})}   \\
    \FinSuc{}_{\V{n}}     & \Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
\end{array}
}
\]

@@

\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}ll}
\FinD\: \V{n} = \DSigma{}{} \!\!\!\!\!\!& (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
                        & \bigRedBracket{
                          \begin{array}{l}
                            \DSigma{\Nat}{\LAM{\V{m}} \DConst{(\V{n} \PropEq \NatSuc{\V{m}})}} \\
                            \DSigma{\Nat}{\LAM{\V{m}} \DProd{\DVar{\V{m}}}{\DConst{(\V{n} \PropEq \NatSuc{\V{m}})}}}
                          \end{array}
                          }
\end{array}
}\]

Intuitively, the constraint is captured as follows. First, we store an
element $\V{m}$ of $\Nat$, thanks to a $\DSigma$. However, the constraint
stipulates that $\V{m}$ cannot be \emph{any} natural numbers: it must be
``the index minus one''. This translates into the constraint $\V{n}
\PropEq \NatSuc{\V{m}}$.

@@

The $\Fin$ example is also a well-known example of Generalised
Abstract Data-Type (GADT) in Haskell. It comes as no surprise that our
universe of indexed description supports this restricted class of
inductive families. As we have seen with $\Fin$, constrained
constructors are straightforwardly translated into equations.

@@

Further, we can actually remove these equations altogether. Let us
look back at $\Fin$. We note that the equations are introduced because
we are \emph{storing} the index of the inductive family, through
$\V{m}$. Hence the constraints. However, \emph{inductive families need not
  store their indices}~\cite{brady:index-inductive-families}. Applying
the \emph{forcing} optimisation to our initial definition of $\Fin$
gives the following, equivalent definition:

\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}llll}
\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{(\EnumT{[]})}{\Void} } \\
\FinD\: (\NatSuc{\V{n}}) & = & \DSigma{}{}\!\!\!\!\!\!\! & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
                     &   &             & \bigRedBracket{
                                         \begin{array}{l}
                                         \DConst{\Unit} \\
                                         \DVar{\V{n}}
                                         \end{array}
                                         }
\end{array}
}\]

@@

\[\stk{
\begin{array}{ll}
\FreeIMonad{\_}{} : & _{\PITEL{\V{I}}{\Set}}
                     \PITEL{\V{R}}{\TagIDesc{\V{I}}} 
                     \PITEL{\V{X}}{\V{I} \To \Set}\To 
                      \TagIDesc{\V{I}}
\end{array} \\
\FreeIMonad{(\V{E},\V{F})}{\V{I}}{\V{R}} \mapsto \\ \;\;
    \pair{\pair{\ListCons{\DVar{}}{(\fst{\V{E}})}} 
               {\LAM{\V{i}}
                \pair{(\DConst{(\V{R}\: \V{i})})}
                     {((\snd{\V{E}})\: \V{i})}{}}{}}
         {\V{F}}{}
}\]

@@

Another instance of free monad is $\IDesc$ itself. Indeed, $\DVar$ is
nothing but the \return. The remaining constructors are simply the
pattern functor, trivially indexed by $\Unit$. Hence, we describe
this functor by the following code:

\[\stk{
\IDescFreeD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescFreeD~\V{I} \mapsto \DSigma{}{} & (\EnumT \red{[} \DConst{},
                                              \DProd{}{},
                                              \DSigma{}{}, 
                                              \DPi{}{} \red{]}) \\
                                  & \bigRedBracket{\begin{array}{l}
                                        \DConst{\Set}               \\
                                        \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
                                    \end{array}}
\end{array}
}\]

>>>>

However, this new class of objects cannot be defined by mere
induction. In the case of vectors, for instance, we have to define the
whole \emph{family} of vector in one go: vectors of all size needs to
be defined at the same time. Where simply-typed languages have
inductive types, in dependently typed programming we the basic grammar of
data-types should be that of inductive families of types. To this end, we
rely on \emph{indexing}. We are going to transform the $\Desc$ universe into
$\IDesc{\!}$, the universe of indexed descriptions. We express inductive families
by the $\V{I} \To \IDesc{\V{I}}$ type.  \note{pwm: Vec might not be the best
motivating example here, since it doesn't appear in the paper (more on that
later). Come to think of it, this is entirely too vague without spelling out
a concrete example.}

@@

Before moving on indexed descriptions, we have to carry out some
maintenance work on descriptions. We presented $\Desc$ as the grammar
of inductive types. Hence, the codes closely follow this grammar. In
the following, we adopt an alternative presentation. With
$\DSigma{\!}{\!}$, we are actually \emph{quoting} a standard
type-former, namely

@@

In the alternative presentation, we go further and present all our
codes as quotations of standard type-formers. This presentation is
shown in Figure~\ref{fig:type-former-desc}. The reader will notice
that we replace $\DUnit$ by a more general $\DConst{\!}$ code. Whereas
$\DUnit$ was interpreted as the unit set, $\DConst{X}$ is interpreted
as $X$, for any $\Bhab{X}{\Set}$. Extensionally, $\DConst{X}$ and
$\DSigma{X}{\DUnit}$ are equivalent. However, $\DConst$ is more
succinct. More importantly, $\DConst$ is \emph{first-order}, unlike
its equivalent encoding. From a definitional perspective, we are
giving more opportunities to the type-system, hence reducing the
burden on the programmer. For the same reason, we introduce
$\DProd{\!}{\!}$ that overlaps with $\DSigma{\!}{\!}$.

This reorganisation is strictly equivalent to the previous one
(Fig.~\ref{fig:hindx_desc}). Just as the previous version, it is also
self-descriptive. We refer the reader to the companion technical
report for details. In this finer-grained presentation, we can define
$\DIndx{\!}$ and $\DHindx{\!}{\!}$ as follow:

@@

\[\stk{
\NatD : \Desc \\
\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{\!} ]})}
                     {[ \DUnit \quad \DId ]}
}\]

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where                                      \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DId            & \Desc                                    \\
    \DConst{}       & \Set \To \Desc                           \\
    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\Desc} \Desc          \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc 
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\descop{\DId}{\V{X}}          &  \V{X}                                           \\
\descop{\DConst{\V{Z}}}{\V{X}}    &  \V{Z}                                           \\
\descop{\DProd{\V{D}}{\V{D'}}}{\V{X}} &  \TIMES{\descop{\V{D}}{\V{X}}}{\descop{\V{D}\V{'}}{\V{X}}}       \\
\descop{\DSigma{\V{S}}{\V{D}}}{\V{X}} &  \SIGMA{\V{s}}{\V{S}} \descop{\V{D}\: \V{s}}{\V{X}}          \\
\descop{\DPi{\V{S}}{\V{D}}}{\V{X}}    &  \PI{\V{s}}{\V{S}} \descop{\V{D}\: \V{s}}{\V{X}}            
\end{array}
}
\end{array}
}\]

@@

In the previous section, we have presented the $\Desc$ universe as the
grammar of functors in the category $\Set$. We have seen how to code
inductive types in this setting. To describe an inductive family
indexed by $\Bhab{\V{I}}{\Set}$, we use endofunctors on the category
$\Set^{\V{I}}$. We call these \emph{indexed functors}. \note{Predicate
  transformer anyone?} In this setting, $\V{I} \To \IDesc{\V{I}}$ is our
grammar for describing these functors. Hence, $\IDesc{\!}$ and its
interpretation have the following types:

\[\stk{
\data \IDesc{\!} (\Bhab{\V{I}}{\Set}) : \Set \where \\
\;\;\ldots \\
\\
\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{\V{I}} \To (\V{I} \To \Set) \To \Set    \\
\ldots
}\]

@@

However, we still have to define the actual grammar. This can be
achieved by a minor evolution of $\Desc$
(Fig.~\ref{fig:type-former-desc}). Indeed, the functorial nature of
$\Desc$ is entirely captured by the $\DId$ code. We get an indexed
functor by turning $\DId$ into an index-dependent $\DVar{\!}$
code. The resulting code of $\IDesc{\!}$ is presented in
Figure~\ref{fig:idesc}.

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \IDesc{\!} (\Bhab{\V{I}}{\Set}) : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DVar{}         & I \To \IDesc{I}                                   \\
    \DConst{}       & \Set \To \IDesc{I}                                \\
    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\IDesc{I}} \IDesc{I}           \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I}  \\
    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I} 
\end{array}
}
\vspace{0.2in}
\\
\stk{
\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{\V{I}} \To \Set) \To \Set                  \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{\V{i}}}{\V{I}}{\V{P}}      &  \V{P}~\V{i}                                                 \\
\idescop{\DConst{\V{X}}}{\V{I}}{\V{P}}    &  \V{X}                                                   \\
\idescop{\DProd{\V{D}}{\V{D'}}}{\V{I}}{\V{P}} &  \TIMES{\idescop{\V{D}}{\V{I}}{\V{P}}}{\idescop{\V{D'}}{\V{I}}{\V{P}}}       \\
\idescop{\DSigma{\V{S}}{\V{D}}}{\V{I}}{\V{P}} &  \SIGMA{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}                    \\
\idescop{\DPi{\V{S}}{\V{D}}}{\V{I}}{\V{P}}    &  \PI{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}            
\end{array}
}
\end{array}
}\]

@@

In the case of vectors, for instance, for the index $\NatZero$, we
would only propose the constructor $\ListNil$. Similarly, for
$\NatSuc{n}$, we only propose the constructor $\ListCons{\!}{\!}$. Hence,
we obtain the following definition:

@@

In order to gain some intuition with $\IDesc{\!}$, let us re-implement
the pattern functor of natural numbers. We refer the reader to
Section~\ref{sec:desc-examples} for its incarnation in the universe of
descriptions. The code is the following:

\[\stk{
\NatD : \IDesc{\Unit} \\
\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{\!} ]})}
                     {[ \DConst{\Unit} \quad (\DVar{\Void}) ]}
}\]

@@

A mandatory exercise consists in describing $\IDesc{\!}$ in itself. With
the experience gained in Section~\ref{sec:desc-levitate}, this
levitation does not pose any issue:

\[\stk{
\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescD~I \mapsto \DSigma{}{} & (\EnumT \red{[} \DVar{},
                                          \DConst{},
                                          \DProd{}{},
                                          \DSigma{}{}, 
                                          \DPi{}{} \red{]}) \\
                              & \bigRedBracket{\begin{array}{l}
                                      \DConst{\V{I}}                  \\
                                      \DConst{\Set}               \\
                                      \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                      \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
                                      \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
                                   \end{array}
                             }
\end{array}
}\]

Note that this description also lives in $\IDesc{\Unit}$. Indeed,
for any $\V{I}$, $\IDesc{\V{I}}$ is merely an inductive type. As for the
$\Desc$ levitation, the finite function space $[ \ldots ]$ elaborates into a
special purpose $\F{switchID}$ operator.

@@

In our universe of descriptions, this data-type is represented by a
tagged indexed description. We use the index to carry the type: the
resulting description is indexed by $\Ty$. We observe that some
constructors are ``polymorph'', namely $\Econd{}{}{}$ and $\Eval{}$. On the
other hand, the $\Eplus$ and $\Ele$ constructors are
index-dependent. $\Eplus$ is defined if and only if the result type --
the index -- is $\Enat$, whereas $\Ele$ is defined if and only if the
index is $\Ebool$. The actual code precisely follows this intuition,
as shown in Figure~\ref{fig:hexpr-full}. For brevity, we use an
informal $\case{\ldots}{\ldots}$ notation, simulating a definition by
pattern-matching. Formally, this corresponds to a call to the
$\switch{\!}{\!}{\!}{\!}$ eliminator.

@@

Another common dependent data-type is $\Fin{\!}$, the finite sets of
numbers. $\Fin{\!}$ is a typical example of an inductive family using
\emph{constrained constructors}. The constructors $\FinZero$ and
$\FinSuc{\!}$ are only defined for an index distinct from $\NatZero$:

\[
\stk{
\data \Fin{\!} : \PI{\V{n}}{\Nat} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FinZero_n      & \Fin{(\NatSuc{n})}   \\
    \FinSuc{}_n     & \Fin{n} \To \Fin{(\NatSuc{n})}
\end{array}
}
\]

@@

\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}ll}
\FinD\: n = \DSigma{}{} & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
                        & \bigRedBracket{
                          \begin{array}{l}
                            \DSigma{\Nat}{\LAM{m} \DConst{(n \PropEq \NatSuc{m})}} \\
                            \DSigma{\Nat}{\LAM{m} \DProd{\DVar{m}}{\DConst{(n \PropEq \NatSuc{m})}}}
                          \end{array}
                          }
\end{array}
}\]

Intuitively, the constraint is captured as follows. First, we store an
element $m$ of $\Nat$, thanks to a $\DSigma$. However, the constraint
stipulates that $m$ cannot be \emph{any} natural numbers: it must be
``the index minus one''. This translates into the constraint $n
\PropEq \NatSuc{m}$.

@@

The $\Fin{\!}$ example is also a well-known example of Generalised
Abstract Data-Type (GADT) in Haskell. It comes as no surprise that our
universe of indexed description supports this restricted class of
inductive families. As we have seen with $\Fin{\!}$, constrained
constructors are straightforwardly translated into equations.

@@

Further, we can actually remove these equations altogether. Let us
look back at $\Fin{\!}$. We note that the equations are introduced because
we are \emph{storing} the index of the inductive family, through
$\V{m}$. Hence the constraints. However, \emph{inductive families need not
  store their indices}~\cite{brady:index-inductive-families}. Applying
the \emph{forcing} optimisation to our initial definition of $\Fin{\!}$
gives the following, equivalent definition:

\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}llll}
\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{\Void}{\Void} } \\
\FinD\: (\NatSuc{n}) & = & \DSigma{}{} & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
                     &   &             & \bigRedBracket{
                                         \begin{array}{l}
                                         \DConst{\Unit} \\
                                         \DVar{n}
                                         \end{array}
                                         }
\end{array}
}\]

@@

\[\stk{
\begin{array}{ll}
\FreeIMonad{\_}{} : & _{\PITEL{\V{I}}{\Set}}
                     \PITEL{\V{R}}{\TagIDesc{\V{I}}} 
                     \PITEL{\V{X}}{\V{I} \To \Set}\To 
                      \TagIDesc{\V{I}}
\end{array} \\
\FreeIMonad{(E,F)}{I}{X} \mapsto
    \pair{\pair{\ListCons{\DVar{}}{(\fst{E})}} 
               {\LAM{i}
                \pair{(\DConst{(X\: i)})}
                     {((\snd{E})\: i)}{}}{}}
         {F}{}
}\]

@@

Another instance of free monad is $\IDesc{\!}$ itself. Indeed,
$\DVar{\!}$ is nothing but the \return. The remaining constructors are
simply the pattern functor, trivially indexed by $\Unit$. Hence, we
describe this functor by the following code:

\[\stk{
\IDescFreeD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescFreeD~I \mapsto \DSigma{}{} & (\EnumT \red{[} \DConst{},
                                              \DProd{}{},
                                              \DSigma{}{}, 
                                              \DPi{}{} \red{]}) \\
                                  & \bigRedBracket{\begin{array}{l}
                                        \DConst{\Set}               \\
                                        \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})} \\
                                        \DSigma{\Set}{(\LAM{\V{S}} \DPi{\V{S}}{(\LAM{\_} \DVar{\Void})})}
                                    \end{array}}
\end{array}
}\]

### 2010-03-29 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: fix spacing of operators used in text.

The type of finite sets comes with further machinery. Indeed, to
manipulate the $\EnumU$ and $\EnumT{}$ objects, we need to equip our
type theory with an appropriate \emph{elimination
  principle}. Formally, as we add types for data -- such as finite
sets here --, we need such primitive operators to compute with those
data, and corresponding extension of the judgemental equality with
their computational behavior. For compactness and readability, we
shall present these operators as functional programs, much as we
implement them in our Agda model. Hence, we first define the
$\spi{}{}$ operator:

\[\stk{
%% spi
\spi{}{} : \PITEL{\V{e}}{\EnumU}
           \PITEL{\V{P}}{\EnumT{\V{e}} \To \Set} \To \Set \\
\begin{array}{@{}ll@{\:\mapsto\:\:}l}
\spi{\NilE}{& \V{P}}        & \Unit \\
\spi{(\ConsE{\V{t}}{\V{e}})}{& \V{P}} & \TIMES{\V{P}\: \Ze}{\spi{\V{e}}{(\LAM{\V{x}} \V{P}\: (\Su{\V{x}}))}}
\end{array}
}\]

A closer look to its type should reveal a certain familiarity with its
equivalent for sets, $\Pi$. Indeed, $\spi{}{}$ defines a \emph{finite
  function space}, mapping elements of finite sets to objects in
$\Set$. Intuitively, this operator build a right-nested tuple,
effectively mapping an object of $P\:i$ for each $i$ the finite
domain.

Having understood the meaning of $\spi{}{}$, the definition of the
$\F{switch}$ eliminator is not surprising:

@@

As we have seen, finite sets are not just a way of making
data. Through the $\spi{}{}$ and $\F{switch}$ operators, they also
come with a notion of finite function space. However, we had to
extract that intuition from the type, by a careful reading. This seems
to contradict our argument in favour of types for coding
representation. Here, we are overflown by low-level, very precise type
information.

>>>>

The type of finite sets comes with further machinery. Indeed, to
manipulate the $\EnumU$ and $\EnumT{\!}$ objects, we need to equip our
type theory with an appropriate \emph{elimination
  principle}. Formally, as we add types for data -- such as finite
sets here --, we need such primitive operators to compute with those
data, and corresponding extension of the judgemental equality with
their computational behavior. For compactness and readability, we
shall present these operators as functional programs, much as we
implement them in our Agda model. Hence, we first define the
$\spi{\!}{\!}$ operator:

\[\stk{
%% spi
\spi{\!}{\!} : \PITEL{\V{e}}{\EnumU}
           \PITEL{\V{P}}{\EnumT{\V{e}} \To \Set} \To \Set \\
\begin{array}{@{}ll@{\:\mapsto\:\:}l}
\spi{\NilE}{& \V{P}}        & \Unit \\
\spi{(\ConsE{\V{t}}{\V{e}})}{& \V{P}} & \TIMES{\V{P}\: \Ze}{\spi{\V{e}}{(\LAM{\V{x}} \V{P}\: (\Su{\V{x}}))}}
\end{array}
}\]

A closer look to its type should reveal a certain familiarity with its
equivalent for sets, $\Pi$. Indeed, $\spi{\!}{\!}$ defines a
\emph{finite function space}, mapping elements of finite sets to
objects in $\Set$. Intuitively, this operator build a right-nested
tuple, effectively mapping an object of $P\:i$ for each $i$ the finite
domain.

Having understood the meaning of $\spi{\!}{\!}$, the definition of the
$\F{switch}$ eliminator is not surprising:

@@

As we have seen, finite sets are not just a way of making
data. Through the $\spi{\!}{\!}$ and $\F{switch}$ operators, they also
come with a notion of finite function space. However, we had to
extract that intuition from the type, by a careful reading. This seems
to contradict our argument in favour of types for coding
representation. Here, we are overflown by low-level, very precise type
information.
